### 2020-07-21 manuscript/v2_medrxiv/discussion.Rmd
remarks twitter part 1

In addition, in alignment with other recent published studies, our results confirm evidence which associated a negative effect of temperature on new COVID-19 cases [@bannister-tyrrell_preliminary_2020;@demongeot_temperature_2020;@liu_impact_2020;@qi_covid-19_2020;@shi_impact_2020;@sobral_association_2020;@tosepu_correlation_2020;@Wang2020temperature;@wu_effects_2020]. It is however controversial to other scientific literature describing no effects [@briz-redon_spatio-temporal_2020;@iqbal_nexus_2020;@jahangiri_sensitivity_2020;@juni_impact_2020;@yao_no_2020] or even converse correlations [@auler_evidence_2020;@xie_association_2020]. The conflicting results might be explained by different climates and characteristics of the populations under study. While we are confident that our strict causal analysis resulted in effect estimates as undistorted as possible, there might be unconsidered bias in those other studies. Further research needs to be done to elucidate the biological characteristics of the novel virus SARS-CoV-2 regarding its ambient temperature survival and transmission. In regards of humidity, our findings (higher humidity leading to increased case numbers) are in agreement with previous research [@qi_covid-19_2020;@wu_effects_2020;@juni_impact_2020]. Finally, we found a positive causal effect of increment precipitation and a raise in COVID-19 cases, which supports previous observations [@sobral_association_2020].

@@

While use of a causal DAG is itself a strong tool to identify causal effects (and not just statistical associations), it introduces two limitations: causal assumptions within the graph (depicted by edges) need to be well justified, and the statistical regression model that calculates total causal effects needs to be appropriate for the task at hand. We endorse our graph as a basis for discussion on residual confounding.

We observed overdispersion and a substantial increase in model performance with a negative binomial regression compared to Poisson regression, which is in line with the results on COVID-19 daily case counts of @kraemer_effect_2020 and others [@liu_impact_2020;@bannister-tyrrell_preliminary_2020]. We did not model case counts with a differential equation model like the classic SIR-model [@kermack1991contributions] and its successors, since these are more suited to prediction [e.g. @rkimodellierung] while our choice of a negative binomial regression framework allowed us to estimate the effects of confounders more reliably. There are more advanced statistical methods for count data, e.g. zero-inflated models and mixed models. We tested both approaches as extensions to the negative binomial regression and experienced numerical problems and increased computing time, along with an insubstantial increase in model performance. Furthermore, our model assumed that all variables have effects proportional to the size of their measurements. It is possible that some variables show saturation effects or opposite effects for lower, medium, or high values. This could be modelled with polynomial or other transformations of the variables, which we did not employ due to limited temporal and spatial data availability.

As we were using administrative data for our analysis, the results are susceptible to the Modifiable Area Unit Problem (MAUP) [@Openshaw1984]. The MAUP postulates that different regional aggregations of the units of observation may lead to different results and conclusions. Due to limited available data for the different variables, there is currently no way to overcome this.

@@

In our analysis, the adverse effect of mobility in retail/recreation and the favorable effect of mobility in grocery/pharmacy make a strong case for the successful implementation of policies like contact restrictions which limit the number of individual interactions that lead to infections. This is due to retail/recreational areas encompassing mostly places of social gatherings like restaurants and bars, malls, sports and music venues, among others, while if people are doing more of their essential shopping at supermarkets, they will most likely stay at home with less contact to other people.

@@

To the best of our knowledge, this is the most comprehensive analysis of causes for COVID-19 infections which integrates different data sources (all publicly available). Causal reasoning with a DAG allows us to estimate the causal effects reliably.

>>>>

In addition, in alignment with other recent published studies, our results confirm evidence which associated a negative effect of temperature on new COVID-19 cases [@bannister-tyrrell_preliminary_2020;@demongeot_temperature_2020;@liu_impact_2020;@qi_covid-19_2020;@shi_impact_2020;@sobral_association_2020;@tosepu_correlation_2020;@Wang2020temperature;@wu_effects_2020]. It is however controversial to other scientific literature describing no effects [@briz-redon_spatio-temporal_2020;@iqbal_nexus_2020;@jahangiri_sensitivity_2020;@juni_impact_2020;@yao_no_2020] or even converse correlations [@auler_evidence_2020;@xie_association_2020]. The conflicting results might be explained by different climates and characteristics of the populations under study. While we are confident that our strict causal analysis resulted in effect estimates as undistorted as possible, there might be unconsidered bias in those other studies. Further research needs to be done to elucidate the biological characteristics of the novel virus SARS-CoV-2 regarding its ambient temperature survival and transmission. Finally, we found a positive causal effect of increment precipitation and a raise in COVID-19 cases, which supports previous observations [@sobral_association_2020].

@@

While use of a causal DAG is itself a strong tool to identify causal effects (and not just statistical associations), it introduces two limitations: causal assumptions within the graph (depicted by edges) need to be well justified, and the statistical regression model that calculates total causal effects needs to be appropriate for the task at hand. We endorse our graph as a basis for discussion on residual confounding. We did not try to construct the DAG from the available data (cf. [@gencoglu2020causal]). As such, our proposed DAG is not entirely consistent with the data and there are conditional dependencies between variables that cannot be dissolved by adding edges to the DAG (e.g. between the policies like contact restrictions and mandatory face masks).

We observed overdispersion and a substantial increase in model performance with a negative binomial regression compared to Poisson regression, which is in line with the results on COVID-19 daily case counts of @kraemer_effect_2020 and others [@liu_impact_2020;@bannister-tyrrell_preliminary_2020]. We did not model case counts with a differential equation model like the classic SIR-model [@kermack1991contributions] and its successors, since these are more suited to prediction [e.g. @rkimodellierung] while our choice of a negative binomial regression framework allowed us to estimate the effects of confounders more reliably. There are more advanced statistical methods for count data, e.g. zero-inflated models and mixed models. We tested both approaches as extensions to the negative binomial regression and experienced numerical problems and increased computing time, along with an insubstantial increase in model performance. Furthermore, our model assumed that all variables have effects proportional to the size of their measurements. It is possible that some variables show saturation effects or opposite effects for lower, medium, or high values. This could be modelled with polynomial or other transformations of the variables, which we did not employ due to limited temporal and spatial data availability. Use of a fixed DAG with effect estimation via regression assumes that data was generated by the same underlying process for the observation period. By inclusion of the successive mitigation policies as binary variables we were able to explain some of the variance caused by the changing dynamics of case numbers, though.

We stress the point that our effects were deduced on an aggregate (district) level in the absence of available data on an individual level. As such, we cannot make conclusions about effects on individuals without the possibility for an ecological fallacy. Furthermore, as we were using administrative data for our analysis, the results are susceptible to the Modifiable Area Unit Problem (MAUP) [@Openshaw1984]. The MAUP postulates that different regional aggregations of the units of observation may lead to different results and conclusions. Due to limited available data for the different variables, there is currently no way to overcome this.

@@

In our analysis, the adverse effect of mobility in retail/recreation and the favorable effect of mobility in grocery/pharmacy indicate that policies like contact restrictions which limit the number of individual interactions can lead to reduced infection numbers. This is due to retail/recreational areas encompassing mostly places of social gatherings like restaurants and bars, malls, sports and music venues, among others, while if people are doing more of their essential shopping at supermarkets, they will most likely stay at home with less contact to other people.

@@

To the best of our knowledge, this is the most comprehensive analysis of causes for COVID-19 infections which integrates different data sources (all publicly available). Causal reasoning with a DAG allows us to estimate the causal effects more reliably.

### 2020-07-21 manuscript/v2_medrxiv/results.Rmd
remarks twitter part 1

Within our framework, we saw strong effects for mobility in retail/recreational areas and essential shopping (grocery and pharmacy). In the former, an increase of 1 percent point mobility compared to the reference period (03 January to 06 February 2020) leads to an increase of the daily reported case number by about `r round(100*abs(1-causaleffects%>%filter(cause=="Mobility (retail and recreation)")%>%pull("effect estimate")), 1)` percent. Contrarily, a corresponding increase of 1 percent point for the areas of grocery/pharmacy leads to a decrease in the reported case number by approximately `r round(100*abs(1-causaleffects%>%filter(cause=="Mobility (grocery and pharmacy)")%>%pull("effect estimate")), 1)` percent. Mobility on workplaces showed a small effect of `r round(100*abs(1-causaleffects%>%filter(cause=="Mobility (workplaces)")%>%pull("effect estimate")), 1)` increase in case numbers for every 1 percent point increase in mobility. Other causal effects of mobility were insubstantial and not consistent in their direction (99 percent confidence intervals of estimates include 1). Figure \@ref(fig:effectsrange) shows the effects of mobility on a range of possible values. Thus, we expect an increase of daily cases by approximately `r round(100*abs(1-(causaleffects%>%filter(cause=="Mobility (retail and recreation)")%>%pull("effect estimate"))^(0-mymeans["Mobility (retail and recreation)"])), 0)` percent if mobility in retail/recreation reaches baseline levels of 0 percent difference to the reference period. On the other hand, an increase of mobility for grocery/pharmacy by 10 percent points compared to the reference period leads to a reduction of the infection rate by approximately `r round(100*abs(1-(causaleffects%>%filter(cause=="Mobility (grocery and pharmacy)")%>%pull("effect estimate"))^(10-mymeans["Mobility (grocery and pharmacy)"])), 0)` percent.

@@

Within our model, we observed a causal effect of temperature and all other weather variables. Every increase of 1 degree Celsius in temperature leads to a reduction of the daily reported case numbers by approximately `r round(100*abs(1-causaleffects%>%filter(cause=="Temperature")%>%pull("effect estimate")), 1)` percent. Similarly strong was the effect of rainfall: One millimeter (=1 liter per square meter) more rainfall leads to an increase of reported case numbers by approximately `r round(100*abs(1-causaleffects%>%filter(cause=="Rainfall")%>%pull("effect estimate")), 1)` percent. We observe small effects for humidity and wind as well (higher humidity leading to fewer cases, stronger wind leading to less cases). In perspective (Figure \@ref(fig:effectsrange)), with temperature we expect an increase by approximately `r round(100*abs(1-(causaleffects%>%filter(cause=="Temperature")%>%pull("effect estimate"))^(0-mymeans["Weather (temperature)"])), 0)` percent at a daily average temperature of $0^{\circ}C$. For rainfall, we expect on a very rainy day with 10 mm rainfall a corresponding increase of the infection rate by approximately `r round(100*abs(1-(causaleffects%>%filter(cause=="Rainfall")%>%pull("effect estimate"))^(0-mymeans["Weather (rainfall)"])), 0)` percent.

>>>>

Within our framework, we saw significant effects for mobility in retail/recreational areas and essential shopping (grocery and pharmacy). In the former, an increase of 1 percent point mobility compared to the reference period (03 January to 06 February 2020) leads to an increase of the daily reported case number by about `r round(100*abs(1-causaleffects%>%filter(cause=="Mobility (retail and recreation)")%>%pull("effect estimate")), 1)` percent. Contrarily, a corresponding increase of 1 percent point for the areas of grocery/pharmacy leads to a decrease in the reported case number by approximately `r round(100*abs(1-causaleffects%>%filter(cause=="Mobility (grocery and pharmacy)")%>%pull("effect estimate")), 1)` percent. Mobility on workplaces showed a small effect of `r round(100*abs(1-causaleffects%>%filter(cause=="Mobility (workplaces)")%>%pull("effect estimate")), 1)` increase in case numbers for every 1 percent point increase in mobility. Other causal effects of mobility were insubstantial and not consistent in their direction (99 percent confidence intervals of estimates include 1). Figure \@ref(fig:effectsrange) shows the effects of mobility on a range of possible values. Thus, we expect an increase of daily cases by approximately `r round(100*abs(1-(causaleffects%>%filter(cause=="Mobility (retail and recreation)")%>%pull("effect estimate"))^(0-mymeans["Mobility (retail and recreation)"])), 0)` percent if mobility in retail/recreation reaches baseline levels of 0 percent difference to the reference period. On the other hand, an increase of mobility for grocery/pharmacy by 10 percent points compared to the reference period leads to a reduction of the infection rate by approximately `r round(100*abs(1-(causaleffects%>%filter(cause=="Mobility (grocery and pharmacy)")%>%pull("effect estimate"))^(10-mymeans["Mobility (grocery and pharmacy)"])), 0)` percent.

@@

Within our model, we observed a causal effect of temperature and all other weather variables. Every increase of 1 degree Celsius in temperature leads to a reduction of the daily reported case numbers by approximately `r round(100*abs(1-causaleffects%>%filter(cause=="Temperature")%>%pull("effect estimate")), 1)` percent. On the other hand, we found an increasing effect of rainfall: One millimeter (=1 liter per square meter) more rainfall leads to an increase of reported case numbers by approximately `r round(100*abs(1-causaleffects%>%filter(cause=="Rainfall")%>%pull("effect estimate")), 1)` percent. We observe small effects for humidity and wind as well (higher humidity leading to fewer cases, stronger wind leading to less cases). In perspective (Figure \@ref(fig:effectsrange)), with temperature we expect an increase by approximately `r round(100*abs(1-(causaleffects%>%filter(cause=="Temperature")%>%pull("effect estimate"))^(0-mymeans["Weather (temperature)"])), 0)` percent at a daily average temperature of $0^{\circ}C$. For rainfall, we expect on a very rainy day with 10 mm rainfall a corresponding increase of the infection rate by approximately `r round(100*abs(1-(causaleffects%>%filter(cause=="Rainfall")%>%pull("effect estimate"))^(0-mymeans["Weather (rainfall)"])), 0)` percent.
